A spiking neural integrator model of the adaptive control of action by the medial prefrontal cortex

### 2013-11-14 paper/rebuttal.tex

\documentclass[11pt,paper=letter]{scrartcl}
\usepackage[english]{babel}
\usepackage{amssymb,amsmath}
\usepackage{namedplus}
\usepackage[font=sffamily]{quoting}
\usepackage{helvet}
\usepackage[backgroundcolor=yellow,linecolor=yellow]{todonotes}
\usepackage{hyperref}
\hypersetup{
  colorlinks = true, %Colours links instead of ugly boxes
  urlcolor = blue, %Colour for external hyperlinks
  linkcolor = blue, %Colour of internal links
  citecolor = blue %Colour of citations
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{12pt}

\begin{document}

\vspace*{1.5\baselineskip}
\begin{flushright}
  \today
\end{flushright}

Dear Editor,

Please find a revised version of
``A spiking neural integrator model of the adaptive
control of action by the medial prefrontal cortex''
attached for your consideration for publication in the
Journal of Neuroscience.

We very much appreciate the reviews
of the original submission.
Considerable effort has been made to address
all issues raised in the reviews.
In order to address these issues,
we have done additional simulations
and revised a significant portion of the text.
Three new figures and several pages
of text have been added
to clarify issues that were unclear
in the original submission,
and to present the results
of the additional simulations.

One notable change in the manuscript,
reflected in the updated title,
is our use of the term ``medial prefrontal cortex,''
or mPFC, instead of ``anterior cingulate cortex,''
or ACC, to characterize the brain area
that was the focus of our study.
This change was made in response to
Reviewer 1, who raised concerns
about functional homologies across species.
Anatomists working on the ACC
consider the prelimbic cortex
(area 32) as a cingulate region
(e.g., studies from the Barbas lab;
\citenoparens{Medalla2009}).
However, there is an unfortunate controversy
about the rat prefrontal cortex,
due in part to the Uylings review
that the reviewer mentioned.
We would like to avoid debate
about this issue in our paper,
and chose to use a more generic term
that is commonly used in rodent studies
to describe the cortical region
in the revised manuscript.

We hope that you find the manuscript
is now suitable for publication
and look forward to hearing from you soon.

Sincerely, \vspace{1em}

Trevor Bekolay, Mark Laubach, Chris Eliasmith

\clearpage
\setlength{\parindent}{1cm}
\setlength{\parskip}{3pt}

\section*{Reviewer 1}

We would like to thank the reviewer for his or her
careful reading of our manuscript
and thoughtful feedback on conceptual
and technical issues in the study.
We have addressed all points raised
in the review,
and we believe that the paper
is very much improved thanks
to Reviewer 1's comments.

\begin{quoting}
  The computational model, named ``double integrator model'' is a
  dynamical system composed of two interacting populations of spiking
  neurons. It is programmed in the Nengo simulation environment, as
  part of the Neural Engineering Framework (NEF; Eliasmith and
  Anderson, 2003) which has the potential to constitute a useful tool
  for neuroscientists to try to simulate, reproduce and better
  understand activities of large cortical networks.

  The description of the model should be strongly clarified and
  extended to enable the reader to precisely understand what is being
  simulated and to distinguish what is part of the general Nengo
  framework and what is specific to the proposed computational model.
\end{quoting}

We appreciate the comments on Nengo and the NEF.
The ``double integrator model'' is originally
from \citetext{Singh2006}.
We have adapted it for the present study
by giving it more inputs, with the most
important being the event that terminates
integration (reward feedback).
Additionally, we have made
a feedforward control network,
and used the double integrator
to control its behavior adaptively.
These distinctions have been made more clear
due to reorganization of the methods section,
and by simplifying several sections of the model.
For example, the \textit{Holding} and \textit{Trigger}
populations only existed in the neural implementation.
We have removed those and ensured that
all neural models are a direct implementation
of their corresponding dynamical systems.
All of the equations governing
the dynamical systems are included
in the revised manuscript.

\begin{quoting}
  More analyses of the results should be provided (dependence to
  parameters, robustness) to better highlight under which conditions
  the model can or cannot reproduce the results - instead of simply
  showing a version of the model that works well.
\end{quoting}

We have now included several numerical simulations
of the dynamical system (Figures 4 and 5)
to clarify the goals of the model,
and under which conditions it meets these goals.
These simulations show how the system changes
as a result of changes to the three free parameters
in the double integrator network equations.
The many parameters affecting the neural implementation
are selected partly at random;
the model performs well and
matches the experimental data
for many different sets of randomized parameters.
We have explicitly listed the distributions
from which we choose neural parameters
in order to make the model easier to reproduce.

\begin{quoting}
  The relation to previous computational models and experimental data
  of the primate ACC and rodent mPFC should be better framed and
  discussed, since it is not clear how the present simulations of one
  particular rat mPFC electrophysiological study can
  relate/complete/replace previous theories of the role of primate ACC
  in cognitive control, as the authors claim in the introduction; the
  discussion should be extended to thoroughly address each raised
  issue.
\end{quoting}

Thanks for this comment.
We must admit that some of the text
in the original manuscript was too bold
with regard to the impact of our model
of delayed response performance
by the rat mPFC.
The text was revised with this in mind,
and we hope that the changes
address Reviewer 1's concerns.

There are many models of ACC/mPFC function
available in the literature.
Given the scope of our study, it was not possible
to address all potential models.
We chose to focus on the recent PRO model
by \citetext{Alexander2011} and included
comparisons of our results
to that model in Section 5.1 and Figure 6.
We have also included Section 5.2 in the discussion
to relate our results to those from
reinforcement learning based approaches like
\citetext{Behrens2007,Alexander2011,Khamassi2012}.

\begin{quoting}
  In the introduction, the objective of the study and its potential
  contribution to important neuroscience debates should be better
  formulated. First, the authors claim that "there are currently no
  mathematical accounts of ACC function that would explain the myriad
  of behaviors modulated by the ACC." But the model presented here
  does not attempt to explain a myriad of behaviors. It only
  reproduces data from a single study (Narayanan and Laubach, 2009).
\end{quoting}

The reviewer is correct, and we should not
have written this text so strongly.
The introduction has been thoroughly
revised in the new manuscript.

\begin{quoting}
  Although I agree that there currently is not any complete model of
  ACC functions, there are several recent computational models that
  have importantly extended the initial models and that the authors do
  not seem aware of (Behrens et al., 2007; Alexander and Brown, 2011;
  Khamassi et al., 2013). I think the manuscript would greatly benefit
  from clearly stating what is new compared to these previous models.
\end{quoting}

As mentioned above, a detailed comparison
to \citetext{Alexander2011} is included
and the other two papers are now cited and discussed.

\begin{quoting}
  Can these models already explain part of the results of (Nayanan and
  Laubach, 2009)? Even if yes, I think that the dynamical aspect of
  the present model provides something more that should be better
  highlighted. For instance, I think that both the models of Alexander
  and Brown 2011 and Khamassi et al. 2013 could account for the ACC
  activity reported in Narayanan and Laubach 2009 that "change
  significantly after errors, and appear to contain information that
  an error occurred". This should be discussed.
\end{quoting}

Reviewer 1 is correct that the models presented in
\citetext{Alexander2011} and \citetext{Khamassi2012}
would be able to account for some elements
of the \citetext{Narayanan2009} data.
However, we believe that our ability
to generate spike data that can be
directly compared to the experimental data
is a unique contribution,
in additional to the dynamical aspect
that the reviewer notes.
While the information
being tracked is similar between
all of these models
(as shown in the revised manuscript's Figure 6),
we interpret that information
from a dynamical systems perspective
which provides an alternative hypothesis
of how that information might come about
(as Reviewer 1 states).
We have highlighted the dynamical aspect of the model
(particularly in Section 5.2, but also throughout),
and thank Reviewer 1 for pointing out
the similarities between these models.

\begin{quoting}
  It is not true that since ``no existing models have been implemented
  with spiking neurons [...] no models can be analyzed with the same
  methods used to analyze single neuron recordings.'' Simulations of
  simple non-spiking neural network models or even of algorithmic
  models can be analyzed with methods that consider neurons' firing
  rate in different time bins.
\end{quoting}

We respectfully disagree on this point.
While it is true that
our analyses could be performed with
firing rates in time bins,
there are more detailed analyses that can only
be performed with raw spike data.
For neural network models with
real-valued output from each neuron,
a representation scheme must be used
to represent that real-valued output
with spiking neurons;
firing rate alone is rarely sufficient.
For rate models that emit firing rates
and use a representation scheme,
the temporal dynamics in a spiking model
are significantly different;
a neural integrator, for example,
requires about twice as many
spiking neurons than rate neurons
to obtain similar performance.

In short, there are a variety
of differences between rate and spiking models
that require additional analyses
or ``translations'' of rate models
for comparison with spike data,
which are not required of spiking models.

\begin{quoting}
  I think the important thing here is that some characteristics of
  recorded neural activities can be seen as emerging properties of a
  dynamical system which a spiking neural network model enables to
  reproduce.
\end{quoting}

We agree with this statement,
and have attempted to highlight this
in the revised discussion.
As a general statement on neural modeling,
we believe that the aim should not be
to reproduce neural activities,
but instead to produce a working mathematical model.
By implementing the model in a spiking neural system,
we ensure that the model is robust
to the temporal dynamics of spiking neural networks,
and are able to compare directly
to recorded neural activities.
In this sense, the neural activities can be thought of
as ``emerging'' from the neural realization
of the mathematical model.

\begin{quoting}
  I think there is a major issue that needs to be solved concerning
  the acceptability or non-acceptability of considering that the mPFC
  (more precisely the prelimbic cortex (PL)) in rodents is the same
  thing as the ACC in primates. The state of the art of related
  previous computational models provided in the introduction only
  discusses models of ACC function in primates. It does not discuss
  any computational model of PL in rodents (e.g. see Koene and
  Hasselmo, 2005; Martinet et al., 2011). Classically, the rodent PL
  is considered as subserving homologous functions to the dorsolateral
  prefrontal cortex (DLPFC) in primates (Uylings et al., 2003). But
  people agree that the PL in rodents is less differentiated than the
  DLPFC in primates, and PL may integrate functions homologous to both
  the primate DLPFC and ACC. However, it should be taken into
  consideration that the more dorsal part of mPFC in rodents (named
  ACd/Fr2 in the Paxinos) could also partly integrate some functions
  homologous to the ACC in primates. Globally the authors should give
  more justifications. Cite for instance Seamans et al. (2008).
  Nevertheless, it is not straightforward to draw conclusions about
  ACC functions in primates from a model that has been calibrated to
  reproduce electrophysiological activities from a single study in the
  rodent PL (i.e. Narayanan and Laubach, 2009). Maybe the paper would
  benefit from a rewriting of the Introduction mainly describing
  previous experimental results and computational models of PL in
  rodents (although still briefly mentioning the implications for the
  understanding of the role of ACC in primates), and discussing in
  details the possible homologies with the ACC in primates in the
  Discussion.
\end{quoting}

The terms used for describing the medial frontal cortex
across species continues to be an issue for us (ML)
and many other researchers.
The original manuscript used
the term ``anterior cingulate cortex,'' or ACC.
Our rationale for this was that anatomists
working on the primate ACC
have commonly included the prelimbic cortex
(area 32) in the ``ACC''
(for example, studies from the Barbas lab;
\citenoparens{Medalla2009}).
However, there is an unfortunate controversy
about the rat prefrontal cortex,
due in part to the Uylings review
that the reviewer mentioned.
One of us (ML) has spent much time
discussing this issue
in conferences and workshops
(see \citenoparens{Laubach2011}).
We would like to avoid debate about this issue
in our paper, as it has little to do
with the validity of a double integrator network
as a candidate model
for explaining computations
performed by the rodent PFC
in controlling timed actions.
We chose to use a more generic term
(medial prefrontal cortex, or mPFC),
which is commonly used in rodent studies,
to describe the cortical region in the revised manuscript.
We have also limited the scope of our claims.

Following the reviewer's suggestion,
we added text to the Introduction
to review the series of studies
done by the Laubach lab
that led to collaboration
with the Eliasmith lab
on the double integrator network.

\begin{quoting}
  The presentation of previous results obtained by Narayanan and
  Laubach 2009 is too short and is not described with enough details
  to enable the reader to fully understand without reading the
  original paper. The authors should dedicate more space to describe
  the results displayed in Figure 2 and to clearly explain the
  interpretation they make. In particular, what does figure 2F tell
  us? Also, we would like to see on figure 2 the onset of the correct
  and error related PCS at the time of the feedback.
\end{quoting}

Thanks for raising this issue.
We agree that more should have been done to make Figure 2 clear,
and panel F was not sufficiently explained
in the original paper.
A paragraph was added to Section 2.5.1
to discuss the plots in panel F,
and the timing of reward feedback
has been noted in the figure.
Importantly, we note that
the analysis was only done on correct trials,
and so feedback was at the same time
for bot the post-correct and post-error trials.

\begin{quoting}
  The model should be described in more details both for specialists
  who would like to reimplement it themselves and for experimentalists
  who will read the paper. First, the terms used in equation (6)
  should be better explained. Second, it is not always clear which
  computations are specific to the proposed computational model and
  which computations are general to the NEF framework to implement any
  dynamical system and ``any neuron model''. To make the message
  clearer, I think it is important here to stick to the description of
  the proposed computational model, and to mention in a clearly
  separated paragraph what describes properties of the NEF (and of
  Nengo) that could be interesting to a reader that would like to
  program his own model.
\end{quoting}

Some confusing and unnecessary aspects of the model
have been simplified.
For example, the \textit{Holding} and \textit{Trigger} populations
were removed.
They originally existed in order to extend
the model to more general situations
in which motor actions may be
contingent on body state
through visual or proprioceptive feedback;
however, we have instead constrained the model
and linked the results only to the simple RT task
in order to simplify explanation
and avoid overgeneralizing the model
to tasks we have not modeled.

We have made several improvements to
the methods sections on
the dynamical systems and the NEF.
Included in those improvements
is a rewritten Equation (6)
that now uses matrix notation
to make the mathematical operation
(least-squares minimization) more clear.
While it was also the case
in the original paper
that NEF-specific equations
were only listed in one section,
we hope that with these improvements
it is more clear that all of the equations
in Section 2.4 are specific to the NEF,
and all equations in other sections
are part of the dynamical systems
that are simulated both directly
and implemented in spiking networks with the NEF.

Due to the lack of supplementary material,
it is difficult to include
a sufficiently detailed explanation of the NEF
to make it easily reproducible
from this paper alone.
We have opted for a comprehensive
explanation that includes
all of the equations necessary
to reproduce the results,
but with shorter explanations
than can be found
in other resources on the NEF.
However, we believe that the open source nature
of Nengo and the scripts that we use
to generate the model
make it possible for specialists
to reproduce the model.

\begin{quoting}
  The authors should avoid using the term "model" for many different
  things, since this can be very confusing for experimentalist readers
  (i.e. the computational model; "any neuron model"; "for each
  instance of each model"; ...).
\end{quoting}

The many levels of modeling that are used in this study
makes an overuse of the word ``model''
difficult to avoid;
we thank the reviewer for pointing this out.
We have attempted to minimize
the use of the word model.
Specifically, we refer to the dynamical systems
(the double integrator, cue-responding,
and adaptive control systems) and their neural
implementations as ``networks.''
All uses of the term ``model'' are prefaced
with a identifying term such as
``neuron'' or ``network'' in order
to avoid ``model'' losing its meaning
as a mathematical idealization.

\begin{quoting}
  The authors mention a mechanism for switching between an aggressive
  timing strategy and a conservative cue-response strategy. What is
  the programmed criterion for switching? Please write the used
  equations.
\end{quoting}

Thanks to Reviewer 1 for pointing out this glaring omission.
We have included the switching criterion as Equation (5)
in the revised manuscript.

\begin{quoting}
  It is important to provide analyses of the strengths and weaknesses
  of the proposed computational model, so that the reader can really
  understand what enables and what does not enable the model to
  succeed in reproducing the experimental data. What is the
  sensibility to changes in parameter values? Could the authors
  illustrate the quality of reproduction of experimental data as a
  function of different parameter sets?
\end{quoting}

We have attempted to explain the meaning of the parameters
through the newly added text in Section 3.1,
and its associated analyses in Figure 4.
We have also given more context to these parameters
in the explanations of the dynamical systems
in Sections 2.2 and 2.3.

Aside from the three parameters we set,
the model's success is dependent solely
on the ability of the integrative populations
to retain their represented values over time.
In our case, model fits are generally good
(Pearson $R^2 > 0.7$)
in any case when the random neural parameters
produce good integrators.
The ALIF neuron introduces additional dynamics
that make creating neural integrators
more difficult than
neural integrators made up of LIF neurons,
but these difficulties can be overcome
with larger populations.
The original double integrator paper
\cite{Singh2006}
discusses the robustness
of NEF-created neural integrators
in the double integrator context,
and \citetext{Eliasmith2005} discusses this issue
for many NEF-constructed attractor networks.
More detailed analyses can be found
in \citetext{Eliasmith2003}.

\begin{quoting}
  Related to this issue, on page 11, an important question is how much
  of parameter tuning was necessary to obtain the same PCs as
  experimental data? It is good to reproduce the experimental results.
  But it is even better to show analyses of different regimes or
  states the model can produce with different parameters, and to
  highlight which of these states or regimes may the experimental data
  correspond to.
\end{quoting}

We now discuss in the paper
how each parameters affects the model,
and show simulations with different sets of parameters
(Table 1 describes the ALIF parameters;
Table 2 and Figure 4 describe the
dynamical system parameters).
Section 3.1 discusses the results
of the simulations
and lists the parameters values
that were chosen.
Importantly, the parameter choices
did not require fitting,
and for most parameters,
a range of values would produce
functionally identical results.

For the experimental model fits in Figure 7,
several models were generated and the best
experimental fit was chosen.
It is possible to show the
average fits for
several instantiations of the model,
but this is not very informative because
increasing the number of neurons in the model
produces networks that fit the data well
nearly all the time,
despite many randomly generated parameters.
As noted above, the performance of the integrators
is the critical factor to the performance
of the entire model.

### 2013-11-14 paper/jneurosci2013.tex
Removed some packages from .tex
Revised version submitted Nov 29
- Responded to JNeurosci reviews
- Now using doit to analyze, plot etc
- Many other improvements; see the diff

\documentclass[11pt]{article}
\usepackage{amsmath,amsfonts,amsthm,amssymb}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{extramarks}
\usepackage{chngpage}
\usepackage{caption}
\usepackage{soul}
\usepackage{namedplus}
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx,float,wrapfig}
\usepackage{listings}
\usepackage{framed}
\usepackage{inconsolata}
\usepackage{hyperref}

\hypersetup{
  colorlinks = true, %Colours links instead of ugly boxes
  urlcolor = blue, %Colour for external hyperlinks
  linkcolor = blue, %Colour of internal links
  citecolor = red %Colour of citations
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Make title
\title{A spiking neural integrator model of the adaptive control of
  action by the anterior cingulate cortex}
\date{Abbreviated title: A spiking model of ACC adaptive control}
\author{%
  Trevor Bekolay \\
  Centre for Theoretical Neuroscience \\
  University of Waterloo \\ ~ \\
  Mark Laubach \\
  The John B. Pierce Lab \\
  New Haven, CT \\ ~ \\
  Chris Eliasmith \\
  Centre for Theoretical Neuroscience \\
  University of Waterloo}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\textbf{Pages:} 24. \textbf{Figures:} 8. \textbf{Tables, multimedia, 3D models:} 0

\textbf{Word counts:} Abstract: 164, Introduction: 445, Discussion: 479

@@

\begin{abstract}
  Reaction times can be improved by preparing movements
  in advance and by learning the expected timing
  of action-imperative stimuli.
  Success or failure on the previous trial
  can be an important factor for determining
  whether an animal will attempt to time
  the stimulus or simply wait for it to occur
  before initiating action.
  The anterior cingulate cortex (ACC) has been implicated
  in enabling the top-down control of action
  depending on the outcome of the previous trial.
  Analysis of ACC activity suggests that
  neural integration is a key mechanism
  for adaptive control in precisely timed tasks.
  We show through simulation
  that a spiking neural circuit
  consisting of coupled neural integrators
  captures the neural dynamics
  of the experimentally recorded ACC.
  We expand on this coupled integrator model
  to construct a spiking neural circuit that
  performs a reaction time task by following
  either a timing or cue-response strategy,
  and show that it performs a simple reaction time task
  with similar reaction times as experimental subjects.
\end{abstract}

@@

The anterior cingulate cortex (ACC) has been implicated
in a wide variety of functions,
including conflict monitoring
\cite{Sheth2012,Aarts2009,Botvinick2004,VanVeen2004},
performance monitoring
\cite{Brown2005,Carter1998,Fecteau2003,Modirrousta2008,Rushworth2004,Histed2009,Hyman2013,Horst2012},
anticipatory control \cite{Aarts2008,Koyama2001},
arousal of the sympathetic nervous system
\cite{Critchley2003,Luu2003},
and control of action timing
\cite{Muir1996,Mulert2003,Naito2013,Narayanan2009,Risterucci2003}.
Conceptual models have been proposed
that would allow the ACC to be involved
in several of these functions,
depending on behavioral context
(e.g., \citenoparens{Botvinick2004,Luu2003}).
However, while there have been
attempts at modeling the ACC
using artificial neural networks
\cite{Botvinick2004,Brown2005},
there are currently no mathematical
accounts of ACC function that would
explain the myriad of behaviors
modulated by the ACC.
As well, no existing models
have been implemented with spiking neurons,
and therefore no models can be analyzed
with the same methods used
to analyze single neuron recordings.

In this paper, we propose that a
two-dimensional dynamical system
can account for ACC activity
related to error monitoring and motor action timing.
We test the model in a simple reaction time task
with a predictable foreperiod and
show that the mathematical model
enables precise timing of actions,
and changes behavior depending
on the outcome of the last trial.
The dynamical system is based
on a multi-dimensional attractor
network proposed by \citetext{Singh2006},
and is implemented in a spiking neural network
using the Neural Engineering Framework
(NEF; \citenoparens{Eliasmith2003}).
We show, using identical analyses
to those employed on the experimental data,
that simulated neural activity from this circuit
closely matches recorded ACC activity
and that the model can be combined with
a simple motor control model
to produce behavior consistent
with behavioral studies of simple reaction time tasks.

Our model also provides an alternative hypothesis
of how the ACC accomplishes the control of action timing.
\citetext{Narayanan2009} hypothesized
that ACC activity inhibits motor responses
until the time of the cue.
Our model instead proposes that
ACC activity encodes the duration of the action
and that inhibition is carried out elsewhere.

In this paper we apply our
mathematical account of ACC function
to explain error monitoring and
the control of action timing.
However, we suggest that the dynamical modelling
method we employ is general and can be extended
to explain other theories
of high-level ACC function.

@@

\subsection{The reaction time task} \label{sec:simple-rt}

Twelve male Long-Evans rats
were trained to press down on a lever
for a fixed amount of time (the ``foreperiod'';
1 second for this study)
using standard methods \cite{Laubach2000}.
After the foreperiod, a cue is presented
(an auditory tone).
The subject then has 0.6 seconds (the response window)
in which to release the lever.
If the lever is released in that time,
the trial is a success, and the subject is rewarded
with water.
If the lever is released during the foreperiod,
the trial is classified as a ``premature error,''
and is penalized by a time-out with the house lights extinguished.
If the lever is not released within the response window,
the trail is classified as a ``late error,''
and is also penalized by a time-out with the house lights extinguished.
Figure~\ref{fig1} presents a schematic view of the task.
Subjects were considered to have learned the task
when they performed in correctly $> 60 \%$
of the time.

The same reaction time task was simulated
in the Nengo simulation environment
using a finite state machine to track
task state, and scalar values between -1 and 1
to indicate the state of house lights,
lever position, auditory cue, and reward.
The Nengo scripts simulating this environment
are available at
\url{https://github.com/tbekolay/jneurosci2013}.

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath,amsfonts,amsthm,amssymb}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{extramarks}
\usepackage{chngpage}
\usepackage{caption}
\usepackage{namedplus}
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphicx,float}
\usepackage{listings}
\usepackage{framed}
\usepackage{inconsolata}
\usepackage{hyperref}
\usepackage{booktabs}
\hypersetup{
  colorlinks = true, %Colours links instead of ugly boxes
  urlcolor = blue, %Colour for external hyperlinks
  linkcolor = blue, %Colour of internal links
  citecolor = blue %Colour of citations
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Make title
\title{A spiking neural integrator model of the adaptive control of
  action by the medial prefrontal cortex}
\date{Abbreviated title: A spiking model of mPFC adaptive control}
\author{%
  Trevor Bekolay \\
  Centre for Theoretical Neuroscience \\
  University of Waterloo \\ ~ \\
  Mark Laubach \\
  The John B. Pierce Lab and Dept. of Neurobiology \\
  Yale University School of Medicine \\ ~ \\
  Chris Eliasmith \\
  Centre for Theoretical Neuroscience \\
  University of Waterloo}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\textbf{Pages:} 32. \textbf{Figures:} 10. \textbf{Tables:} 2. \textbf{Multimedia, 3D models:} 0

\textbf{Word counts:} Abstract: 199, Introduction: 486, Discussion: 1499.

@@

\begin{abstract}
  Subjects performing simple reaction time tasks
  can improve reaction times by learning the expected timing
  of action-imperative stimuli
  and preparing movements in advance.
  Success or failure on the previous trial
  is often an important factor for determining
  whether a subject will attempt to time
  the stimulus or wait for it to occur
  before initiating action.
  The medial prefrontal cortex (mPFC) has been implicated
  in enabling the top-down control of action
  depending on the outcome of the previous trial.
  Analysis of spike activity from the rat mPFC
  suggests that neural integration is a key mechanism
  for adaptive control in precisely timed tasks.
  We show through simulation
  that a spiking neural network
  consisting of coupled neural integrators
  captures the neural dynamics
  of the experimentally recorded mPFC.
  Errors lead to deviations in the normal
  dynamics of the system,
  a process that could enable learning
  from past mistakes.
  We expand on this coupled integrator network
  to construct a spiking neural network that
  performs a reaction time task by following
  either a cue-response or timing strategy,
  and show that it performs the task
  with similar reaction times as experimental subjects
  while maintaining the same spiking dynamics
  as the experimentally recorded mPFC.
\end{abstract}

@@

The medial prefrontal cortex (mPFC) has been implicated in regulating reinforcement learning parameters
\cite{Amiez2005,Khamassi2012},
conflict monitoring
\cite{Botvinick2004,VanVeen2004,Aarts2009,Sheth2012},
performance monitoring
\cite{Carter1998,Fecteau2003,Rushworth2004,Brown2005,Modirrousta2008,Histed2009,Alexander2011,Horst2012,Hyman2013},
anticipatory control \cite{Koyama2001,Aarts2008},
arousal of the sympathetic nervous system
\cite{Critchley2003,Luu2003},
and control of action timing
\cite{Muir1996,Mulert2003,Risterucci2003,Narayanan2009,Naito2013}.
Conceptual models have been proposed
that would allow the mPFC to be involved
in several of these functions,
depending on behavioral context
(e.g., \citenoparens{Luu2003,Botvinick2004}).
Mathematical models
based on reinforcement learning
explain experimental results
at several levels
(e.g., \citenoparens{Alexander2011,Khamassi2012}).
However, these reinforcement learning models
have not been shown to explain spiking activity
of neural ensembles recorded during
temporally constrained behavioral tasks.
Here, we present a dynamical spiking neural model
that relates directly to
both neural and behavioral data
in a simple mPFC-dependent task.

Specifically, we propose a
two-dimensional dynamical system
that can account for mPFC activity
related to error monitoring and action timing
in rodents performing a simple reaction time task.
Neural recordings in the motor cortex
during the acquisition of this task
reveal that an encoding of forthcoming errors
develops with learning \cite{Laubach2000}.
These activities depend on processing in the mPFC
\cite{Narayanan2006Neuron},
where neurons encode errors
both prospectively and retrospectively
\cite{Narayanan2006Neuron,Narayanan2008}.
Analyses of population activity during the task
show that both the mPFC and motor cortex
are dominated by
relatively slow fluctuations in firing rates,
starting from the initiation of the trial
and terminating when feedback is given
about the trial's outcome
\cite{Narayanan2009}.

Here, we show that a specific type of network structure,
called a double integrator network \cite{Singh2006},
can play a central role in
accounting for reaction time task performance.
With analyses identical
to the experimental study,
we show that the neural activity produced
by the simulated spiking network closely matches
the experimental data from the
\citetext{Narayanan2009} study.
Furthermore, we propose a feedforward control system
that can perform the simple reaction time task
and show that the double integrator network
can be used to modify the behavior
of the control system based on
the outcome of the previous trial.
When the control system is implemented
in a spiking neural network,
the modified behavior closely matches
the behavior of experimental subjects
performing the same reaction time task.

@@

\subsection{Reaction time task} \label{sec:simple-rt}

The experimental data analyzed in this study is the same
as that in \citetext{Narayanan2009},
where more detailed explanation of the experimental
methods can be found.
In that study, twelve male Long-Evans rats
were trained to press down a lever
for a fixed amount of time (the ``foreperiod'';
1 second for this study)
using standard methods \cite{Laubach2000}.
After the foreperiod, an auditory cue is presented.
The subject then has 0.6 seconds (the response window)
in which to release the lever.
If the lever is released in that time,
the trial is a success, and the subject is rewarded
with water.
If the lever is released during the foreperiod,
the trial is classified as a ``premature error,''
and is penalized by a time-out with the house lights extinguished.
If the lever is not released within the response window,
the trial is classified as a ``late error,''
and is also penalized by a time-out with the house lights extinguished.
Figure~\ref{fig1}A presents a schematic view of the task.

Rats included in the study reached performance criterion
of $> 60 \%$ correct trials in 9.75 $\pm$ 1.5 sessions
(1748 $\pm$ 343 trials).
Microwire electrode arrays were then implanted
into the prelimbic region of the rat cerebral cortex,
which is generally considered to be
part of the medial prefrontal cortex
\cite{Laubach2011} and the anterior cingulate cortex
in particular \cite{Medalla2009}.
Data were recorded using
a Plexon Many Neuron Acquisition Program
for 3 sessions (567 $\pm$ 98 trials).

\subsubsection{Simulation}

In order to provide appropriate
environmental input for
the networks described
in subsequent sections,
the same reaction time task is simulated
using a finite state machine.
The task can be in one of seven states~($S$):
trial start~(TS), foreperiod~(FP),
timeout~(TO), cue~(C), response window~(R),
reward~(Rw) and intertrial interval~(ITI).
State transitions can be seen in
the graphical depiction of
the finite state machine in Figure~\ref{fig1}C.

The auditory cue, house lights, and reward signals
are provided by the environment
based on the current task state.
\begin{equation} \label{eq:env}
  u_C =
  \begin{cases}
    1 &\text{if } S = \text{C} \\
    0 &\text{otherwise}
  \end{cases}
  \quad
  u_{TO} =
  \begin{cases}
    1 &\text{if } S = \text{TO} \\
    0 &\text{otherwise}
  \end{cases}
  \quad
  u_{Rw} =
  \begin{cases}
    1 &\text{if } S = \text{Rw} \\
    0 &\text{otherwise}
  \end{cases}
\end{equation}
The lever position does not depend on task state,
but instead on press ($u$) and release ($u_r$)
signals provided by a control system
(see section~\ref{sec:control}).
\begin{equation} \label{eq:lever}
  L = \int [-u(t) + u_r(t)] \, dt
\end{equation}

\subsection{Double integrator network} \label{sec:dint}

A key finding of the \citetext{Narayanan2009}
study was that the temporal profiles
of the leading principal component
from the mPFC resembled the cumulative sum
of the second component, and vice versa.
This finding suggested that population activity
in the mPFC might reflect a process of integration.
Using the Nengo simulator \cite{BekolayInpress},
a number of candidate networks
were investigated and one particular network
was found to resemble the dynamics
of the mPFC recordings.
This network is called a ``double integrator network,''
and was first proposed in the context of working memory
by \citetext{Singh2006}.
We hypothesize that the double integrator network
proposed by \citeauthor{Singh2006}
will exhibit the same neural dynamics
as the experimental mPFC
when modified to respond
to events occurring during the simple RT task.

In order to test this hypothesis,
we extend the double integrator network
by providing it with multiple inputs
that represent events occurring during
the simple RT task
(including those in Equation~[\ref{eq:env}]),
while maintaining the same underlying dynamics.
By driving the double integrator network
to certain parts of the state space,
we make it possible to predict
the state of the system
when the cue occurs during the
reaction time task.

The underlying dynamics of the double integrator network
are captured in the following equations:
\begin{align*}
  \dot{x}_1 &= u \\
  \dot{x}_2 &= \beta x_1,
\end{align*}
where $u$ represents an input signal
and $\beta$ is a scalar parameter representing the strength of
the connection between the two integrators.
This results in a two-dimensional system
in which $x_1$ integrates its input,
maintaining the accumulated value over time, and
$x_2$ integrates the value of $x_1$.
This is the system proposed by \citetext{Singh2006},
who predicted that this dynamical system is capable
of time tracking;
because $x_2$ integrates $x_1$,
the amount of time elapsed since
$x_1$ changes can be determined
based on how much $x_2$ has changed.

In \citetext{Singh2006}, neural recordings
were analyzed during a single trial
of the vibrotactile discrimination task.
This corresponds to a single trial
of the simple RT task
in which the input, $u$,
is the lever press,
and the initial state, $x$, is $(0, 0)$.
However, in the simple RT task we perform
many consecutive trials.
We therefore use other inputs from
the environment to match
the trajectories tracked by
the experimental principal components
(shown in Figure~\ref{fig2}).

Since the principal components
resemble the double integrator network
in the post-correct case,
we aim to reset the system to the origin point
after correct trials.
Figure~\ref{fig2}D indicates that
the principal components return to their
original state after approximately 2 seconds,
which is the time at which
the subject receives reward.
Therefore, we hypothesize that
the trial's outcome (i.e., reward delivery
or a timeout) resets the system
to some particular state,
as proposed in \citetext{Narayanan2009}.
In the case of reward delivery,
we implement this hypothesis
by using the reward delivery signal, $u_{Rw}$,
to drive the system toward $(0, 0)$.
\begin{align*}
  \dot{x}_1 &= u - R \, u_{Rw} \, x_1 \\
  \dot{x}_2 &= \beta x_1 - R \, u_{Rw} \, x_2,
\end{align*}
where $R$ is a scalar parameter
indicating how strongly the system
is driven to the origin point.

In the post-error case,
the principal components
in Figure~\ref{fig2}D are
not at the origin point.
Since the lever press increases $x_1$,
when $x_1 > 0$ we can conclude that
the subject is currently performing the task.
We therefore hypothesize that $x_1 < 0$
indicates that an error has recently occurred.
We implement this hypothesis
by using the house lights signal,
$u_{TO}$, to decrease $x_1$.
\begin{equation} \label{eq:dint}
  \begin{aligned}
    \dot{x}_1 &= u - R \, u_{Rw} \, x_1 - E u_{TO} \\
    \dot{x}_2 &= \beta x_1 - R \, u_{Rw} \, x_2,
  \end{aligned}
\end{equation}
where $E$ is a scalar parameter
controlling how quickly $x_1$ decreases.
Note that $x_2$ is not directly driven by $u_{TO}$;
however, because of the $\beta x_1$ term,
$x_2$ will also decrease.
